---
title: "Macroeconometrics and Machine Learning : Lasso method"
author: "Kilian ANDRU, Gaëlle DORMION"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import data

```{r}
library(MTS)
library(fbi)
library(glmnet)
```

```{r}
# Setting start and end date of the analysis
start_date = "1960-01-01"
end_date = "2019-12-01"         # We might want to leave out the covid period


# Load the most recent fred-md dataset

# Load transformed data 
data = fredmd("https://files.stlouisfed.org/files/htdocs/fred-md/monthly/current.csv", 
              date_start = as.Date(start_date), date_end = as.Date(end_date))


var_desc = fredmd_description   

# Period
dates <- data$date
```




## Data formatting

We remove useless columns and columns with NA values
```{r}
#Remove the first column "date"
raw_data <- data[,seq(2,length(data))]

series <- names(data)

#Remove columns with NA values
names(raw_data[, colSums(is.na(raw_data)) > 0])
x <- raw_data[, colSums(is.na(raw_data)) == 0]

```

```{r}
#We check if there are still missing values
table(sapply(x, function(x) sum(length(which(is.na(x))))))
```
We plot the two time series of interest 

```{r}
plot(x$DPCERA3M086SBEA,type = "l",ylab = var_desc$description[var_desc$fred == "DPCERA3M086SBEA"])
```

```{r}
plot(x$RPI, type = "l",ylab = var_desc$description[var_desc$fred == "RPI"])
```

# Let standardize the data
```{r}
x_standard <- x

for (col in names(x)){
  mean <- mean(x[,col])
  st_d <- sd(x[,col])
  x_standard[,col] <- (x[,col] - mean)/ st_d
}

#Time variable
time <- c(1:nrow(x_standard))/12 + 1960

#Number of period
nb_T <- nrow(x_standard)

#Number of variable
nb_N <- ncol(x_standard)

#Plot the series
MTSplot(x_standard,time)
```
The standardized data looks similar as what we got in class

## Lasso regression on real personnal consumption


function for lasso regression
```{r}
Lasso_function <- function(p, r, alpha, variable, plot,residuals){
  if (p <= (r)){
    print("Must have p > r")
  }

  #outcome
  Y <- x_standard[(p+1):nb_T,which(names(x_standard)==variable)]
  #initialize endogenous variable
  X_st <- x_standard[p:(nb_T-1),which(names(x_standard)==variable)]
  if (p >= 2){
    for (i in seq(1,p-1)){
      Y_lag <- x_standard[seq(p-i,nb_T-1-i),which(names(x_standard)==variable)]
      X_st <- cbind(X_st,Y_lag)
    }
  }
  #exogenous varaibles and lag
  for (i in seq(0,r)){
    temp <- x_standard[(p-i):(nb_T-i-1),-which(names(x_standard)==variable)]
    X_st <- cbind(X_st,temp)
  }
  X_st <- as.matrix(X_st)
  
  
  #We use a cross validation to get the optimal lambda
  cv_fit <- cv.glmnet(X_st,Y,alpha=alpha)
  lambda <- cv_fit$lambda.min
  
  #To get the residuals
  if (residuals){
    resid <- Y - predict(cv_fit, newx = X_st)
    return(resid)
  }
  
  #We get the estimated lasso model
  lasso_fit <- glmnet(X_st,Y,alpha=alpha,lambda = lambda)
  #Plot if needed
  if (plot){
    plot(glmnet(X_st,Y,alpha=alpha))
    plot(cv_fit)
  }
  
  return(lasso_fit)
}
```



We can now extract the coefficients to try to forecast the series, we start by getting the coefficients and the corresponding lag
```{r}
list_coeff <- function(Lasso_fit){
  coefficients <- Lasso_fit$beta

  #For Y ariable
  coef_Y <- coefficients[1:13]
  lag <- (1:13)
  coefficients_Y <- data.frame(coef_Y,lag)
  #Remove varaibles with null coeff
  coefficients_Y <- coefficients_Y[coefficients_Y$coef_Y !=0,]

  #For exogenous varaibles
  coef_X <- coefficients[seq(14,length(coefficients))]
  FRED <- row.names(Lasso_fit$beta)[seq(14,length(coefficients))]
  
  #We initialize the dataframe, there is 131 variables for each lag
  coefficients_X <- data.frame(coeff_X = coef_X[seq(1,131)])
  coefficients_X$lag <- 1
  coefficients_X$FRED <- FRED[seq(1,131)]
  #Remove varaibles with null coeff
  coefficients_X <- coefficients_X[coefficients_X$coeff_X !=0,]
  
  #For other lag
  for (i in (1:11)){
    coeff_X <- coef_X[seq(i*131+1,(i+1)*131)]
    temp <- data.frame(coeff_X)
    temp$lag <- i+1
    temp$FRED <- FRED[seq(i*131+1,(i+1)*131)]
    
    #Remove varaibles with null coeff
    temp <- temp[temp$coeff_X != 0,]
    
    #Add to dataframe
    coefficients_X <- rbind(coefficients_X, temp)
  }
  
  return (list(COEF_Y = coefficients_Y, COEF_X = coefficients_X))
}
```




## Forecasting

Here we compute the standardized forecasts
```{r}
forecasting <- function(Lasso_fit, variable, h, standard){

    #outcome
    Y <- x_standard[,which(names(x_standard)==variable)]
    #other variables
    X_st <- x_standard[,-which(names(x_standard)==variable)]
    
    list_coef <- list_coeff(Lasso_fit)
    
    #We initialize the list of forecast values
    forecasts <- c()
    #for each period
    for (i in seq(1,h)){
      value <- 0
      
      #for the endogenous variables, each lag
      for (k in seq(1,nrow(list_coef$COEF_Y))){
        lag <- as.numeric(list_coef$COEF_Y$lag[k])
        #coeff
        coeff <- as.numeric(list_coef$COEF_Y$coef_Y[k])
        
        #if we reuse a forecasted value
        if (i > lag){
          value <- value + forecasts[length(forecasts)-lag+1]*coeff
        
        }else{ #When we use value from the data set
          value <- value + Y[nb_T+ i -lag]*coeff
        
        }
        
      }
      #For endogenous variables
      for (k in seq(1,nrow(list_coef$COEF_X))){
        lag <-list_coef$COEF_X$lag[k]
        #coeff
        coeff <- as.numeric(list_coef$COEF_X$coeff_X[k])
        index <- which(names(X_st)== list_coef$COEF_X$FRED[k])
        
        #There is no forecasted value for exogenous, we only use the value it is in the dataset
        if (i <= lag){
          value <- value + X_st[nb_T+ i -lag, index]*coeff
        }
      }
      forecasts <- c(forecasts,value)
    }
  #If we want standardized data
  if (standard) {
    return(forecasts)
  }else{
    my <- mean(x[,which(names(x)==variable)])
    sy <- sd(x[,which(names(x)==variable)])
    return(forecasts*sy + my)
  }
  
}

```

Forecasting error using naive forecasting

```{r}
forecasting_error <- function(Lasso_fit, variable, h,sigma){
  Forecasts <- forecasting(Lasso_fit, variable, h, standard = FALSE)
  
  min <- c()
  max <- c()
  for (i in seq(1,h)){
    error <- sd(x[,which(names(x)==variable)])*sigma*(i)^(1/2)
    min <- c(min,Forecasts[1] - error)
    max <- c(max,Forecasts[1] + error)
  }
  
  return(data.frame(Forecasts,min,max))
}

```


Plot the forecasted values

```{r}
library(ggplot2)
plot_forecast <- function(forecasts_data, variable, start){
  
  Forecasts <- x[,which(names(x)==variable)]
  
  #forecast horizon
  h <- nrow(forecasts_data)
  
  #time variable
  time <- c(1:(length(Forecasts)+h))/12 + 1960
  
  #DAta to plot
  data_plot <- data.frame(Forecasts)
  data_plot$min <- NA
  data_plot$max <- NA
  
  #Adding forecast
  data_plot <- rbind(data_plot,forecasts_data)
  
  data_plot$time <- time
  
  data_plot <- data_plot[data_plot$time >= start,]
  
  p <- ggplot(data=data_plot, aes(time, Forecasts)) + 
    geom_ribbon(aes(ymin = min, ymax = max), fill = "steelblue2") +
    geom_line(color = "firebrick", size = 0.1)
  
  return(p)
  
}
```
## Application to variables

# Real personal consumption

Now we try to find the best Lasso-VAR model to fit real personal consumption.

We take p and r large, useless variables will be removed with the lasso.
```{r}
p = 13
r =12
alpha = 1
variable = "DPCERA3M086SBEA"


Lasso_fit <- Lasso_function(p = p, r = r, alpha = alpha, variable = variable, plot = TRUE, residuals = FALSE)

print(Lasso_fit$lambda)
print((log(nb_N-1)/nb_T)^(1/2))
print(Lasso_fit$dev.ratio)
```


We extract the residuals to compute the forecasting error
```{r}
residual <- Lasso_function(p = p, r = r, alpha = alpha, variable = variable, plot = FALSE, residuals = TRUE)
sigma <- sd(residual)
```
Number of variables kept, for the outcome
```{r}
coefficients_Y <- Lasso_fit$beta[1:13]
length(coefficients_Y[coefficients_Y !=0])

```

Number of variables kept, for the exogenous variables
```{r}
coefficients_X <- Lasso_fit$beta[seq(14,length(Lasso_fit$beta))]
length(coefficients_X[coefficients_X !=0])
```
We get the varaibles with higher coefficient in absolute value (X_st correspond to Y_lag of order 1)
```{r}
VAR_names <- row.names(Lasso_fit$beta)
coeffs <- Lasso_fit$beta[seq(1,length(Lasso_fit$beta))]
table_coefficients <- data.frame(VAR_names, coeffs)

lags <- c(1,2,3,4,5,6,7,8,9,10,11,12,13)
for (i in seq(1,13)){
  for (k in seq(1,121)){
    lags <- c(lags,i)
  }
}

table_coefficients$lags <- lags
table_coefficients$abs_coff <- abs(table_coefficients$coeffs)

table_coefficients <-table_coefficients[order(table_coefficients$abs_coff,decreasing = TRUE),]

head(table_coefficients)

```


We plot the forecasted values
```{r}
forecast_data <- forecasting_error(Lasso_fit, variable, h= 15,sigma)
plot_forecast(forecast_data, variable, start=1960)

```
We focus on a smaller period
```{r}
plot_forecast(forecast_data, variable, start=2010)
```

# Real personal income

Now we try on real personal income

We take p and r large, useless variables will be removed with the lasso
```{r}
p = 13
r =12
alpha = 1
variable = "RPI"


Lasso_fit <- Lasso_function(p = p, r = r, alpha = alpha, variable = variable, plot = TRUE, residuals = FALSE)

print(Lasso_fit$lambda)
print((log(nb_N-1)/nb_T)^(1/2))
print(Lasso_fit$dev.ratio)
```

We extract the residuals to compute the forecasting error
```{r}
residual <- Lasso_function(p = p, r = r, alpha = alpha, variable = variable, plot = FALSE, residuals = TRUE)
sigma <- sd(residual)
```


Number of variables kept, for the outcome
```{r}
coefficients_Y <- Lasso_fit$beta[1:13]
length(coefficients_Y[coefficients_Y !=0])
```

Number of variables kept, for the exogenous variables
```{r}
coefficients_X <- Lasso_fit$beta[seq(14,length(Lasso_fit$beta))]
length(coefficients_X[coefficients_X !=0])
```
We get the varaibles with higher coefficient in absolute value (X_st correspond to Y_lag of order 1)
```{r}
VAR_names <- row.names(Lasso_fit$beta)
coeffs <- Lasso_fit$beta[seq(1,length(Lasso_fit$beta))]
table_coefficients <- data.frame(VAR_names, coeffs)

lags <- c(1,2,3,4,5,6,7,8,9,10,11,12,13)
for (i in seq(1,13)){
  for (k in seq(1,121)){
    lags <- c(lags,i)
  }
}

table_coefficients$lags <- lags
table_coefficients$abs_coff <- abs(table_coefficients$coeffs)

table_coefficients <-table_coefficients[order(table_coefficients$abs_coff,decreasing = TRUE),]

head(table_coefficients)

```



We plot the forecasted values
```{r}
forecast_data <- forecasting_error(Lasso_fit, variable, h= 15,sigma)
plot_forecast(forecast_data, variable, start=1960)

```

We focus on a smaller period
```{r}
plot_forecast(forecast_data, variable, start=2010)
```
